Magnetic Inhibition of Convection and the Fundamental Properties of Low--Mass Stars

### 2016-03-29 paper/ms.tex
headway on magnetic topology

{\bf The transition exhibited by theoretical magnetic models can be explained by a rapid thinning of outer convective envelopes in stars with masses above about 1.2 $M_{\odot}$ \citep[e.g.,][]{Iben1965}. Once a star reaches an effective temperature of 6\,000 K, it has a very small or non-existent convective envelope, independent of whether or not magnetic inhibition of convection is considered. In the absence of a convective envelope, magnetic inhibition of convection is irrelevant. Therefore, any differences between magnetic and standard stellar model isochrones above this effective temperature are the result of the magnetic field's influence on stellar structure at earlier evolutionary phases (i.e., delayed contraction). This naturally explains the convergence of the 10 Myr magnetic isochrone with the 10 Myr standard model isochrone, which occurs near $6\,000$~K in Fig.~\ref{fig:ages}(b). 
It then seems reasonable to posit that the shallow slope of the empirical HR diagram traces the demise of surface convection zones, the weakening of surface magnetic dynamo action, and the growth of radiative cores in young, pre-main-sequence stars.}

{\bf % Not only for stars with M ~ 1.2... all intermediate mass stars above 0.28 Msun. 
Stars with masses above about $0.28\ M_{\odot}$ are predicted to, at least temporarily, form a small radiative core during their pre-main-sequence contraction. Release of gravitational potential energy as stars contract heats the gas causing radiative opacities to decrease and igniting fusion from the proton-proton ($p$-$p$) chain \citep{Henyey1955}. Stars with masses below about $1.2\ M_{\odot}$ continue to contract until energy production from the $p$-$p$ chain is sufficient to stabilize the core (and thus the star) against further contraction. However, above $1.2\ M_{\odot}$, core temperatures become sufficiently hot enough to produce hydrogen burning via the CN-cycle at an appreciable level. Increased energy production from the CN-cycle, while initially insufficient to support the core against gravitational contraction, causes the radiative core to continue growing and the convective envelope to continue shrinking. Increasing energy production from the CN-cycle and $p$-$p$ chain causes a star's outer layers are predicted to expand while the core's contraction slows. Eventually, the CN-cycle reaches equilibrium, a convective core develops, the radius expansion is reversed, and the convective envelope disappears entirely \citep{Iben1965,Bodenheimer1965}. }

{\bf 
To fully understand the model-predicted transition in Fig.~\ref{fig:ages}(a) and the reasons why magnetic models do not entirely reproduce the morphology of the empirical transition, it is important to consider how the magnetic field evolves as a star's convective envelope shrinks.
Prior to ignition of the CN-cycle, a thinning of the convective envelope means magnetic fields of a given strength will have a larger impact on the thermal structure of the super-adiabatic region. However, a restructuring of the super-adiabatic layers in a star with a thin convective envelope has less of an impact on the thermal structure of the deeper underlying layers, which are assumed to be in radiative equilibrium, than for a star whose interior is largely convective. The thinning of the convective envelope is also accompanied by a drop in the surface gas pressure as the star's radius inflates thereby lowering the permitted equipartition magnetic field strength. The combination of these two effects suggests that the transition from a magnetic sequence to a non-magnetic sequence should coincide with the disappearance of the convective envelope.} % confirm discussion of super-adiabatic layers

{\bf Assuming that a change in slope of the empirical HR diagram is observed at 4\,500 K, 
Comparison of the empirical HR diagram to magnetic stellar models suggests that the transition's effective temperature domain begins somewhere between 4\,500~K and 5\,000~K, consistent with the knee observed in the empirical data at \teff$\sim 4\,500$~K. According to magnetic stellar models, these effective temperatures correspond to stars with masses between 1.26 \msun\ and 1.40 \msun. This is significant because stars in this mass regime are expected to have relatively thin convective envelopes that eventually disappear entirely. 
}

{\bf While this behavior is expected in the magnetic models, it does not need to necessarily correspond to the observed transition. General agreement between the behavior of the magnetic model isochrone and the observed stellar locus provides evidence that magnetic inhibition of convection is the mechanism responsible for the age discrepancy between the high- and low-mass stellar populations. However, it is clear that this statement only holds with respect to the plausible identification of a general mechanism as the models fail to reproduce the precise morphology at the magnetic sequence turn-off.}

{\bf The shallower slope may be an artifact of how the empirical locus was computed. The empirical HR diagram is somewhat under-populated in the immediate vicinity of \teff~$= 5\,500$~K, raising the possibility that the running median has artificially created a smooth transition between warmer and cooler stars. However, the HR diagram is well-populated at slightly warmer and slightly cooler temperatures, suggesting the two regions must be somehow connected. While the precise morphology of the HR diagram in this region may be subject to alteration with additional data, the general trend of should be robust against errors in how the empirical locus is computed.}

@@

If the age difference between warmer and cooler stars in Upper Scorpius is the result of magnetic inhibition of convection, the transition region between $5\,000 <$ \teff/K $< 6\,000$ provides an excellent laboratory for studying magnetic dynamo evolution as a function of convection zone depth and convective core size.

>>>>

{\bf The transition exhibited by theoretical magnetic models can be explained by a rapid thinning of outer convective envelopes in stars with masses above about 1.2 $M_{\odot}$ \citep[e.g.,][]{Iben1965}. Once a star reaches an effective temperature of 6\,000 K, it has a very small or non-existent convective envelope, as shown in Fig.~\ref{fig:rad_core}(b). Figures~\ref{fig:rad_core}(a) and (b) show this is independent of age and mass, a fact arising as a result of the hydrogen ionization zone becoming increasingly localized near the optical photosphere with increasing effective temperature. In the absence of a convective envelope, magnetic inhibition of convection is irrelevant. Therefore, any differences observed between magnetic and standard stellar model isochrones at this effective temperature are the result of the magnetic field's influence on stellar structure at earlier evolutionary phases (i.e., delayed contraction). This provides a natural explanation for the convergence of the 10 Myr magnetic and standard model isochrones that occurs near this effective temperature in Fig.~\ref{fig:ages}(b). 
It then seems reasonable to posit that the shallow slope of the empirical HR diagram and the observed transition in the magnetic models trace the demise of outer convective envelopes and the growth of radiative cores in young, pre-main-sequence stars.}

{\bf While the demise of outer convection zones explains the final convergence of magnetic and standard models, it doesn't necessarily explain the beginning of the transition between  4\,500~K and 5\,000~K. Figures~\ref{fig:ages}(a) and \ref{fig:rad_core}(b) indicate that stars with \teff~<~4\,500~K can possess radiative cores that comprise up to about 75\% of the star, by mass.}

{\bf In the transition region, magnetic models have masses between 1.26 $M_{\odot}$ and 1.40 $M_{\odot}$. Magnetic models predict that these stars have radiative core masses between 76\% and 88\% of their total stellar mass, respectively (see Figure~\ref{fig:rad_core}a). The 1.40 $M_{\odot}$ model is of particular note, as Figure~\ref{fig:mrd} suggests that 1.40 $M_{\odot}$ stars are beginning to enter a brief period of radius inflation that accompanies the ignition of CN-cycle burning \citep{Iben1965}. Photospheric gas pressures rapidly drop during this phase of evolution bringing about a rapid drop in surface equipartition magnetic field strengths. As expected, a rapid decrease in equipartition magnetic field strengths is observed among the data in Table~\ref{tab:equip_values} beginning around 1.40 $M_{\odot}$. Stars above 1.40 $M_{\odot}$ then become less sensitive to magnetic inhibition of convection because their magnetic fields are intrinsically weaker.}

%{\bf Disagreement between model predictions and the properties observed for F-type stars raises concerns about the validity of model predictions during this brief radius inflation phase. Possible reasons for these errors are discussed in Sect.~\ref{sec:radius}. If it }

{\bf If the transition region begins around 4\,500~K, as one might expect from the knee in the empirical HR diagram (Fig.~\ref{fig:empirical}), a rapid decrease in surface magnetic field strengths cannot fully explain the start of the transition. Another mechanism must be responsible and the failure of magnetic models to reproduce the sharp transition suggests the mechanism is not accounted for in the models.}

{\bf The sharp knee at \teff~=~4\,500~K in Fig.~\ref{fig:empirical} may be the result of a global change in magnetic field topology. Evidence for this comes from \citet{Gregory2012}, who noted that large-scale magnetic field topologies observed on pre-main-sequence stars appear to correlate with their predicted radiative core mass. Specifically, they noted a shift from predominantly dipolar axisymmetric magnetic field topologies to multi-polar non-axisymmetric field topologies when the estimated radiative core mass $M_{\rm rad.\ core} > 0.4 \star$. It was noted above that at \teff~=~4\,500~K, magnetic models predict a star with $M_{\star} = 1.26 M_{\odot}$ and $M_{\rm rad.\ core} \approx 0.76 M_{\star}$, well above the predicted radiative core mass where stars are expected to shift toward multi-polar non-axisymmetric field topologies.}

{\bf However, \citet{Gregory2012} used non-magnetic models \citep{Siess2000} to establish a relationship between magnetic field topology and stellar interior structure. Hypothetically, if they were to observe a star in Upper Scorpius with \teff~=~4\,500~K, their HR diagram analysis would suggest that the star is approximately 5 Myr old. Figures~\ref{fig:rad_core}(a) and (b) reveal that a star with \teff~=~4\,500~K at 5 Myr is expected to have a mass of $M_{\star} = 1.10 M_{\odot}$ with $M_{\rm rad.\ core} \approx 0.43 M_{\star}$. This radiative core mass corresponds to 39\% of the total stellar mass and is consistent with the boundary identified by \citet{Gregory2012}.}

{\bf There are several implications of this result. First, the sudden knee observed in Fig.~\ref{fig:empirical} may be the result of a change in magnetic field topology. Second, the shift in magnetic field topology from predominantly axisymmetric to non-axisymmetric may when the radiative core mass is much larger than initially suggested, $M_{\rm rad.\ core} / M_{\star} \approx 0.75$ compared to $M_{\rm rad.\ core} / M_{\star} \approx 0.40$, respectively. Third, pushing the transition in magnetic field topology to higher radiative core masses may provide an explanation for anomalous rotation periods observed among old main-sequence field stars \citep{vanSaders2016}.}

{\bf If the large-scale magnetic field topology of pre-main-sequence stars shifts from predominantly axisymmetric to non-axisymmetric as the radiative core grows beyond 75\% of the total stellar mass, it is conceivable that the reverse is also true. That is, as convective envelopes grow along the sub-giant branch, large-scale magnetic fields may shift from predominantly multi-polar non-axisymmetric toward dipolar axisymmetric topologies. The shift in magnetic field topology may affect magnetic wind breaking mechanisms causing stars to exhibit anomalous rotation periods at older ages \citep{vanSaders2016}. This hypothesis may be at odds with the dependence of magnetic wind breaking mechanisms on large-scale magnetic field topologies, which suggest dipolar magnetic fields exert a greater torque on a star than multi-polar fields \citep[e.g.,][]{Reville2015}. However, it is unclear how the transition from a non-axisymmetric to axisymmetric magnetic field topologies affects this picture.}

{\bf This discussion is premised on the reality of the apparent transition at \teff~=~4\,500~K and the demise of convective envelopes producing the shallower slope in the empirical HR diagram. The shallower slope may be an artifact of how the empirical locus was computed. The empirical HR diagram is somewhat under-populated in the immediate vicinity of \teff~$= 5\,500$~K, raising the possibility that the running median has artificially created a smooth transition between warmer and cooler stars. However, the HR diagram is well-populated at slightly warmer and slightly cooler temperatures, suggesting the two regions must be somehow connected. While the precise morphology of the HR diagram in this region may be subject to alteration with additional data, the general trend of should be robust against errors in how the empirical locus is computed.}

@@

{\bf The observed knee and shallower slope in Fig.~\ref{fig:empirical} appear to be real features. General agreement between the behavior of the magnetic model isochrone and the observed stellar locus provides evidence that magnetic inhibition of convection is the mechanism that is producing these effects. This means that magnetic fields may be responsible for the age discrepancy between the high- and low-mass stellar populations in young associations. However, it is clear that this statement only holds with respect to the plausible identification of a general mechanism as the models fail to reproduce the precise morphology at the magnetic sequence turn-off, perhaps because magnetic models do not account for shifts in global magnetic field topology.}
If the age difference between warmer and cooler stars in Upper Scorpius is the result of magnetic inhibition of convection, the transition region between $5\,000 <$ \teff/K $< 6\,000$ provides an excellent laboratory for studying magnetic dynamo evolution as a function of convection zone depth and convective core size.
